Efficiency

### 2018-06-05 DraftExec.Rnw
with ross acknowledged :-)
latest, with typos for senate

\usepackage{draftwatermark}
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1.5}
\SetWatermarkLightness{0.95}

@@

@@

\begin{titlepage}
    \begin{center}
        \vspace*{1cm}

        \vspace{10cm}
        \textbf{\huge A comparison of pumpjets and propellers for non-nuclear submarine propulsion}

        \vspace{0.5cm}

        \vspace{1.5cm}

        \textbf{Aidan Morrison}

        \vspace{1.5cm}
        January 2018

        \vfill

    \end{center}
\end{titlepage}

@@

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an ``integral part of our bid" \parencite{guillou2017} and that it can be efficient ``across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs that might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

The development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.

@@

However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.  Scientific discovery and technical tends to confirm many of the best discoveries are actually built-upon and confirmed, rather than simply sweep them aside  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.  My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outining the foundational principals.

@@

The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.  The act of slowing the water down is necessarily equivalent to the act of raising its pressure.  It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.  Maintaining this advantage to the same level necessitates a duct design which slows water down to the same extent.  The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.

@@

The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and their known characteristics \ref{buckingham2008submarine}.  A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.  The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.  This advantage reached a proportional maximum at around 6kt of over 60\%.  However, in nominal terms, the greatest differences occurred at slightly lower speeds.  Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.

@@

The full range of plausible efficiency curves exhibited a significant efficiency advantage right through transit speeds.  Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.  None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.  Given the significant emphasis that has been repeatedly placed the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about suitability of this technology to meet Australia's needs.

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

@@

\subsection{They physical relationship}
Perhaps the most important relationship to comprehend in the operation of submarines is the relationship between speed and drag .  This is important because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.  (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction to movement.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging shape) there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

@@

\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different apples called Stokes Equation, in the case where Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}

@@

Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be even quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.  This has considerable consequences for the design of all of the submarine systems, including the propulsion system.

@@

\newpage  \section{Basics of Ducted Propellers and Pumpjets}
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.  Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The difference between a pumpjet and a propeller]
In it simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.  The impact of the shroud in a pumpjet has the special effect of \textit{decelerating} the water at the point that it reaches the propeller.  This has the dual effect of raising the pressure of the water at the point that the rotor has to do its work on it, which is how pumpjets are able to reduce or eliminate cavitation.
\\
\\
This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or `Kort Nozzles'.  This type of duct performs essentially the opposite function to the decelerating duct, contracting in advance of the propeller and acclerating the flow at the stage that it meets propeller.  This effect accentuates the normal contraction of the water-column that occurs in open propellers, and can improve the efficiency of small propellers attempting to exert a large force at low speeds (such as tugs and trawlers) where an optimally sized propeller for such a role would be impractically large for the vessel.  In this circumstance the suction force on the inside of the duct helps generate a positive thrust that is greater than the negative drag on the exterior of the duct.
\end{tcolorbox}

@@

\subsection{Pump Types} \label{pumptypes}
There are broadly three categories of pump types for water which are used for propulsion.  The first is called an axial-flow. In such a pump, the water has work done on in by blades which push the flow in the same direction to the axis of the blades rotation.  This is what a decelerating ducted propeller comprises. The blades of the rotor push the water from the front, to the back, in the same direction as the craft moves, by blades that turn about an axis that is also oriented in that direction.  These types of pumps are generally best for handling very large flows of water, with relatively less head (energy added) per amount of water flowing through, and are consequently generally desired for propulsion systems, where moving water at high velocity, rather than elevating it to high pressure, is the end objective of the system.

@@

\subsection{Recent Commentary}
There has been already some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies on pumpjet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pumpjet for a submarine might perform in quantitative terms, relative to a propeller.

There has also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.  Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.  To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.  In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.  It is noteworthy that some of oldest works, such as those by Wislicenus and Henderson, are still amongst the most frequently cited in the most recent of papers.

@@

Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.  They don't just accelerate a water colum.  They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.  When the total amount of energy embodied in the incoming water stream, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.  The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.  In essence, if you try to slow down something that's already going slowly, it tends to stop  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.  When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.

@@

It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.  The consequences for this pumpjets are and inevitable, since the geometry of the duct is inevitably fixed.  At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.  This adverse pressure gradient is larger at higher speeds, the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is consistent (though still turbulent) throughout the entire cross-sectional area of the duct.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= What about hotel load and the system as a whole?]
The non-propulsive power demand of the submarine (hotel load) has an enormous impact on a conventional submarine's performance underwater. As extra power is drained as time ticks by, underwater endurance and range decline substantially with higher hotel loads.
\\
\\
Hotel load also has a significant impact on the \textit{difference} between the pumpjet and propeller affect the performance of the whole system. By driving the speed at which the propulsion becomes the dominant power demand higher, an extra-large hotel load reduces the overall difference between the two identical submarines, and pushes the peak proportional difference to higher speeds, where endurance and range is naturally lower due to drag increasing.
\\
\\
The inverse relationship leads to even more striking results, particularly if large energy stores (AIP or lithium-ion batteries) are available.  If the hotel load is very low (either because it was designed that way, or discretionary components are shut-down in a high-threat sitation to maximise endurance) then the penalty inflicted by the pumpjet becomes extremely large, quite probably close to a factor of two, amounting to around or over a week of endurance at 3kt, and many hundreds of nautical miles at slightly higher speeds.
\\
\\
The necessary consequence of this dynamic is that selecting a power-hungry combat system such as the AN/BYG-1 in conjunction with the pumpjet and lead-acid batteries for energy when dived will virtually ensure that many substantially smaller conventional sumbarines operating in the Asia Pacific in the coming decades will easily outclass the Shortfin Barracuda in terms of dived range and endurance.
\end{tcolorbox}

@@

\subsection{Indiscretion Ratios}
Indiscretion Ratios represent the fraction of time that a submarine must spend on the surface snorting (running diesel engines) in order to maintain a constant speed.  In order to calculate the Indiscretion Ratio, it is necessary to know how much power the diesel engines are able to generate when running while snorting, and how much energy (net of any efficiency losses, hotel load, and power drawn to maintain propulsion) can be directed to recharging the batteries. Reliable information regarding the arrangement of engines, generators is not generally available in the public domain.  However, from \cite{patrick2012}, it can be seen that credible submarine designs might include diesel engines that have a total capacity in the region of 85\% of the maximum power output of the main motor.  Assuming that there are some necessary losses in the generation and recharging process, and plausibly some increase in the hotel load for the operation of additional systems, fur the purposes of this assessment it is assumed that while snorting approximately 5MW might be available to meet propulsion and normal hotel load requirements, and any surplus can be directed towards charging batteries.

>>>>

% for cover page
\usepackage{geometry}
\usepackage{xcolor}
\usepackage{graphicx}

% \usepackage{draftwatermark}
% \SetWatermarkText{DRAFT}
% \SetWatermarkScale{1.5}
% \SetWatermarkLightness{0.95}

@@

% \SweaveOpts{concordance=TRUE}
\definecolor{titlepagecolor}{cmyk}{0, 0, 0, .30}
\definecolor{namecolor}{cmyk}{0, 0, 0, .80}

@@

%-----------------------------------------------------------------

% ----------------------------------------------------------------
\begin{titlepage}
\newgeometry{left=7.5cm} %defines the geometry for the titlepage
\pagecolor{titlepagecolor}
\noindent

{\huge \textsf{A comparison of pumpjets and}}
\vskip\baselineskip
{\huge \textsf{propellers for non-nuclear}}
\vskip\baselineskip
{\huge \textsf{submarine propulsion}}
\vskip\baselineskip
\noindent
\textbf{Aidan Morrison}
\vskip\baselineskip
\textsf{January 2018}
\vskip\baselineskip
\vskip\baselineskip

\vfill
\noindent

\includegraphics[width=6cm]{tl_logo.png}\\[-1em]
\color{white}
\makebox[0pt][l]{\rule{1.3\textwidth}{1pt}}
\par
\noindent
\textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}

\end{titlepage}
\restoregeometry % restores the geometry
\nopagecolor% Use this to restore the color pages to white

@@

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst the considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one was confident enough on the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an ``integral part of our bid" \parencite{guillou2017} and that it can be efficient ``across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found this seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices ``there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs that might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of ``decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance and consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

The development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.  In addition, the support of my colleague from Trendlock, Ross Ireland, was invaluable in producing the extensive modelling required.

@@

However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.  Scientific discovery and technical progress tends to confirm and build upon many of the best discoveries, rather than simply sweep them aside.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.  My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outining the foundational principals.

@@

The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.  The act of slowing the water down is necessarily equivalent to the act of raising its pressure.  It is this elevated pressure at the point where the blades do their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.  Maintaining this advantage to the same level necessitates a duct design which slows water down to the same extent.  The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.

@@

The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and their known characteristics \parencite{buckingham2008submarine}.  A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.  The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.  This advantage reached a proportional maximum at around 6kt of over 60\%.  However, in nominal terms, the greatest differences occurred at slightly lower speeds.  Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.

@@

The full range of plausible efficiency curves exhibited a significant efficiency advantage right through to transit speeds.  Central assumptions suggested that this would result in the overall range of a submarine being enhanced by approximately 50\% with the use of propellers instead of a pumpjet.  None of the scenarios suggested it was likely that this range advantage would be any less than 20\%.  Given the significant emphasis that has been repeatedly placed on the unique requirements for range and endurance for Australia's submarine, this raises substantial questions about the suitability of this technology to meet Australia's needs.

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

@@

\subsection{The physical relationship}
Perhaps the most important relationship to comprehend in the operation of submarines is the relationship between speed and drag .  This is important because it sets out the fundamental framework as it applies to any submarine, regardless of propulsion type.  (It's also a pretty important for planes, cars, missiles, torpedoes, and basically everything else.)

Drag is the resistance that a fluid (air or water, in our case) gives to a body that is passing through it. Quite simply, it's a force that acts in the opposite direction to movement.  There are multiple sources of drag for different types of scenarios.  For scenarios where an object is in contact with two different types of fluid (like a ship, on the ocean) or when the fluid doesn't really have contact with all of the object (supersonic flight, and supercavitating torpedoes) some more complex physics applies. For a fuller discussion of types of drag in the case of ships, see \cite{carlton2007}.  But in the case of a submarine, which does it's business completely immersed in the ocean, the relevant physics is dominated by the skin friction on the hull, which follows a very simple rule and relationship.  The amount of drag ($F_D$) an objet experiences increases directly in proportion to the surface area $A$.  For any given object of a certain (unchanging) shape there will be a constant coefficent ($C_D$) which reflects how aerodynamic or hydrodymanic the shape is.  The drag is also directly proportional to the density of the fluid being moved through, $\rho$.  (Air creates roughly one-thousandth the drag as water does on any given object at a given speed, since it's roughly one thousand times less dense.)

@@

\footnote{Technical aside: This law applies wherever the the flow over the surface is turbulent. It is true that for very small objects, or very viscous fluids, or very slow movements a different rule applies called Stokes Equation, in the case where Reynolds numbers are less than 1.  Given that sea-water is not particularly viscous, and sumbarines not particularly small, Reynolds numbers are liekly to be much much greater than 1 (one or two thousand), even when moving at only one or two knots.  Since it is unlikely that a significant proportion of the flow over the hull will be laminar, we'll use the drag equation in all modeling going forward when considering drag on the hull.}

@@

Consequently, the difference between the power output a conventional and a nuclear submarine might regularly operate at could be quite a bit larger than the maximum amount of energy that they can deliver to their propulsion systems.  This has considerable consequences for the design of all of the submarine systems, including the propulsion system.

@@

\newpage  \section{Basics of Ducted Propellers and Pumpjets}
It's easy to be somewhat confused by the many different names which seem to be associated with related, or similar systems.  Between pumps, pumpjets, water-jets, ducted or shrouded propellers, or impellers, there is plenty of grounds for some confusion.  In this section, I aim to clarify in simple terms what the important differences between distinctly different systems are, and where some terms are used somewhat interchangeably without doing any great violence to the concepts underlying.  Appreciating where the fundamental as opposed to superficial differences lie is important to understanding subsequently the limitations of different systems.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= The difference between a pumpjet and a propeller]
In its simplest form, a pumpjet can be thought of as a special type of ducted propeller, or propeller contained inside a shroud.  The impact of the shroud in a pumpjet has the special effect of \textit{decelerating} the water at the point that it reaches the propeller.  This has the dual effect of raising the pressure of the water at the point that the rotor has to do its work on it, which is how pumpjets are able to reduce or eliminate cavitation.
\\
\\
This type of duct shape contrasts with the system most commonly associated with ducted propellers on tug-boats and trawlers, known as accelerating ducts or `Kort Nozzles'.  This type of duct performs essentially the opposite function to the decelerating duct, contracting in advance of the propeller and acclerating the flow at the stage that it meets propeller.  This effect accentuates the normal contraction of the water-column that occurs in open propellers, and can improve the efficiency of small propellers attempting to exert a large force at low speeds (such as tugs and trawlers) where an optimally sized propeller for such a role would be impractically large for the vessel.  In this circumstance the suction force on the inside of the duct helps generate a positive thrust that is greater than the negative drag on the exterior of the duct.
\end{tcolorbox}

@@

\subsection{Pump Types} \label{pumptypes}
There are broadly three categories of pump types for water which are used for propulsion.  The first is called an axial-flow pump. In such a pump, the water has work done on in by blades which push the flow in the same direction to the axis of the blades rotation.  This is what a decelerating ducted propeller comprises. The blades of the rotor push the water from the front, to the back, in the same direction as the craft moves, by blades that turn about an axis that is also oriented in that direction.  These types of pumps are generally best for handling very large flows of water, with relatively less head (energy added) per amount of water flowing through, and are consequently generally desired for propulsion systems, where moving water at high velocity, rather than elevating it to high pressure, is the end objective of the system.

@@

\subsection{Recent Commentary}
There has already been some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies on pumpjet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pumpjet for a submarine might perform in quantitative terms, relative to a propeller.

There have also been questions put in parliamentary hearings regarding the pump jet \parencite{patrick2017}, and responses from public officials to such questions indicate that some of the literature provided to support such lines of questions was dated, and that the state of the art may have advanced considerably since this time.  Whilst the statement that the state of the art is much advanced is obviously true in the case with marine propulsion as with most sciences, it is also equally true that new advances tend to confirm, rather than annul, many of the works that have been previously undertaken.  To the extent that modern literature still supports the overall constraints and bounds provided by some of the foundational earlier works, a great deal of useful information can still be obtained from earlier works.  In particular, the foundational theories, often derived from well-established principals of conservation of momentum, energy, as applied by the likes of Bernoulli and Euler, are as well respected and deeply depended upon today as they were centuries earlier when they were first arrived at.  It is noteworthy that some of oldest works, such as those by Wislicenus and Henderson, are still amongst the most frequently cited in the most recent of papers.

@@

Pumpjets are necessarily low in efficiency at very low speeds because they attempt to get the water to do multiple things.  They don't just accelerate a water colum.  They first slow it down (increasing its pressure), then add pressure to it, then accelerate it.  When the total amount of energy embodied in the incoming water stream, and the total amount of energy you want to add to the water both get very small (as they do at very low speed) there simply isn't enough energy to do all things at once to all of the water.  The energy penalty inflicted by using the additional devices becomes very large in proportion, and part of the water doesn't do all the necessary things for the jet to work.  In essence, if you try to slow down something that's already going slowly, it tends to stop.  That's exactly what happens in jets operating at very low flow-rates, when flow separation, also known as stall when it occurs between the blades, begins to take place.  When there isn't enough energy to move all the water through the pressure-hurdles imposed by the duct, only some of it moves, and the rest spins in place.

@@

It also means that for any given change in velocity, the static pressure changes also, but doesn't change as much as the velocity change.  The consequences for pumpjets are inevitable, since the geometry of the duct is inevitably fixed.  At very high flow rates, the expansion of the area ahead of the impeller creates an adverse pressure gradient, as the water must slow down to fill the greater area.  This adverse pressure gradient is larger at higher speeds, but the proportions aren't linear, hence the the speed is much much larger, due to the square relationship, and flow is still consistent (though also turbulent) throughout the entire cross-sectional area of the duct.

@@

\begin{tcolorbox}[width=0.8\textwidth, center, colback=blue!5!white,colframe=blue!75!black, title= What about hotel load and the system as a whole?]
The non-propulsive power demand of the submarine (hotel load) has an enormous impact on a conventional submarine's performance underwater. As extra power is drained as time ticks by, underwater endurance and range decline substantially with higher hotel loads.
\\
\\
Hotel load also has a significant impact on the \textit{difference} between the pumpjet and propeller performance when considering the whole system. By driving the speed at which the propulsion becomes the dominant power demand higher, an extra-large hotel load reduces the overall difference between the two identical submarines, and pushes the peak proportional difference to higher speeds, where endurance and range is naturally lower due to drag increasing.
\\
\\
The inverse relationship leads to even more striking results, particularly if large energy stores (AIP or lithium-ion batteries) are available.  If the hotel load is very low (either because it was designed that way, or discretionary components are shut-down in a high-threat sitation to maximise endurance) then the penalty inflicted by the pumpjet becomes extremely large, quite probably close to a factor of two, amounting to around or over a week of endurance at 3kt, and many hundreds of nautical miles at slightly higher speeds.
\\
\\
The necessary consequence of this dynamic is that selecting a power-hungry combat system such as the AN/BYG-1 in conjunction with the pumpjet and lead-acid batteries for energy when dived will virtually ensure that many substantially smaller conventional sumbarines operating in the Asia Pacific in the coming decades will easily outclass the Shortfin Barracuda in terms of dived range and endurance.
\end{tcolorbox}

@@

\subsection{Indiscretion Ratios}
Indiscretion Ratios represent the fraction of time that a submarine must spend on the surface snorting (running diesel engines) in order to maintain a constant speed.  In order to calculate the Indiscretion Ratio, it is necessary to know how much power the diesel engines are able to generate when running while snorting, and how much energy (net of any efficiency losses, hotel load, and power drawn to maintain propulsion) can be directed to recharging the batteries. Reliable information regarding the arrangement of engines, generators is not generally available in the public domain.  However, from \cite{patrick2012}, it can be seen that credible submarine designs might include diesel engines that have a total capacity in the region of 85\% of the maximum power output of the main motor.  Assuming that there are some necessary losses in the generation and recharging process, and plausibly some increase in the hotel load for the operation of additional systems, for the purposes of this assessment it is assumed that while snorting approximately 5MW might be available to meet propulsion and normal hotel load requirements, and any surplus can be directed towards charging batteries.
